---
title: "Super-Rugby-Try-Analysis"
author: "Andrew Mellor"
date: "23/04/2020"
output: 
  prettydoc::html_pretty:
    theme: cayman
    bibliography: references.bib
    
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Overview

The purpose of this document is to provide analysis about the playing patterns that lead to try scoring in the 2017 Super Rugby season.

## Load required packages

```{r packages, message=FALSE}
library(tidyverse) # include all required packages at the start
```

## Data

The data we will use for this activity is stored in the 2017_super-rugby_try-source-data.csv file. This data consists of tries that were scored during the 2017 Super Rugby competition (observations/rows). 

This data was collected by a former University of Canberra student, Molly Coughlan, as part of a project that identified playing patterns that led to tries in super rugby. 
The [data](https://unicanberra.instructure.com/courses/7493/files/1747546) was provided by Dr Jocelyn Mara from the University of Canberra.


## Read data

Read in the `2017_super-rugby_try-source-data.csv` file using the `read_csv()` function from the `readr` package. 

```{r read_data, message=FALSE} 
# read in the data
# data dir should be in the working dir
df <- read_csv("data/2017_super-rugby_try-source-data.csv")
```

## Checking the data

Check the structure of the data file using `str()`:

```{r structure, message=FALSE}
str(df)
```

Check the first 6 rows of the data file using `head()`

```{r head, message=FALSE}
head(df)
```

Check the last 6 rows of the data file using `tail()`

```{r tail, message=FALSE}
tail(df)
```

Check for missing values using: sum()

```{r na, message=FALSE}
sum(is.na(df)) # count the missing values
head(which(is.na(df), arr.ind = TRUE))
```


## Exploratory analysis

### Try Scoring by Teams

Analysing which teams score more tries than others.
````{r bar}
df%>%
ggplot() +
geom_bar(mapping = aes(x = attacking_team), stat = "count", fill = "blue", colour = "black") +
theme(axis.text.x = element_text(angle = 45, hjust = 1))
````

### Try Scoring by Game Time

### Try Scoring by Play

### Try Scoring by Field Area

### Try Scoring by Time in Possession

### Try Scoring by +/-

## References